Stability of variational quantum dynamics with Gaussian approximants

### 2013-07-29 paper.tex
cleanup for Hui

\nonstopmode
\input a4
\input xpmath
\input respnotes

@@

% It follows that $\G$ is onto: let $|ψ〉$ any ket, and set $f₀(α*)=π^{-½}e^{½|α|²}〈α|ψ〉$.  Then
% $$\G f₀={1\over π}\int |α〉〈α|ψ〉\,d²α=|ψ〉.$$
% Furthermore, coherent states have the property that $e^{½|α|²}|α〉$ is an entire function of $α$ (clarify), so $f₀$ is an entire function.  Glauber showed that it is the unique entire solution to $\G f₀=|ψ〉$.

Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.  It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.  The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.

If, following Bargmann, we identify the quantum space with the entire functions, $|Γ〉$ becomes the projector $〈 Γ⁺|Γ〉$.  Therefore the singularity of $|Γ〉$ is easily resolved.  If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be most simply done by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.  And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.  So it is curious that the discrete version of this problem causes so much trouble.

@@

The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
$$π^{-½}〈β|Γ〉f=π^{-1}∫〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}∫〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.  There are many theorems to the effect that singular values of real convolution operators decay rapidly, and the operators are badly conditioned.  It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.  I think the loophole is that of real functions can only oscillate at a limited rate, so that their Fourier transforms have finite norm.  Complex functions have space in the Argand plane to oscillate arbitrarily rapidy.  And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.

@@

The operator $|Γ〉$ acts on functions defined over the whole complex plane.  For states with high particle number, the smallest preimages have support far from the origin.  We can not compute with such functions: any discretisation will sample values of $f$ on a finite subset of the plane.  Unfortunately, the conditioning of these restricted operators will be worse than that of $|Γ〉$.  We have excluded the small preimages of the states with large particle number, so they must be represented by functions with support close to the origin, but with large norms.  Hence the restricted operators will have small singular values corresponding to states with many particles.

@@

The sizes of states and their expansions can be read from the singular value expansion of $\Gx $.   If the domain and range of $\G$ were the same space, it would have an eigenvalue decomposition; the SVE is a generalisation of that.  It has the form
$$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
The kets $|g_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.  The $σ_i$ are positive real numbers, called singular values.  The functions $v_i$ are are orthonormal under the Bargmann inner product, and are called right singular vectors.

In the case where $χ=1$, and $\Gx=\G$, the expansion is quite arbitrary.  The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire functions for which $\G v_i=|u_i〉$, and every singular value is 1.  If the domain and range of $\G$ were the same, this would indicate that it was a projector.  The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component keeps the same expansion coefficients, while the basis $v_i$ is replaced by the $|u_i〉$.

When the operator is restricted, the SVE gives more information. If $\Gx$ were one to one, each $v_i$ would be the unique preimage of $|g_i〉$; since $\Gx$ is singular, $v_i$ is the preimage with least norm.  The singular values are the ratios of $‖\Gx v_i‖$ to $‖v_i‖$.  If $|u_i〉$ has a low particle number, it can be expanded efficiently: $v_i$ will be close to $〈Γ⁺|u_i〉$, and $σ_i$ will be close to 1.  If $|u_i〉$ includes components with large particle number, its expansion will have a larger norm than itself.  In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, and $σ_i$ will be small.

Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.  Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.  In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.  When the expansion is discretised, however, it is very easy for the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contains the important information about $|ψ〉$ to disappear in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.

Let's compute some singular values and vectors.  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
where
$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
Presumably, the right singular vectors have the $f_n$ of least norm such that 
$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$

@@

\topinsert \XeTeXpicfile rsv2.pdf width 0.5\hsize
\XeTeXpicfile rsv3.pdf width 0.5\hsize \endinsert
\topinsert \XeTeXpicfile rsv4.pdf width 0.5\hsize \endinsert

@@

The most important issue is the Picard condition.  This restates a very basic observation: for the problem $G\dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $f$ over an orthonormal basis must have a finite norm.  If the basis is chosen as the right singular vectors $v_j$, this means that the sum of the squared moduli of the components must be finite.  For $\dot f$ to satisfy the equation, the component of $v_j$ is $〈g_j|H|ψ〉/σ_j$.  So the Picard condition is that, for the equation to have a solution, the series
$$∑_{j=1}^∞{\bigl|〈g_j|H|ψ〉\bigr|²\over σ_j²}$$
must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.

The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.  I was working in a truncated Fock space; I wonder what happens as the truncation increases, while the grid stays a constant size?  Or alternately, if the grid shrinks while the Fock space stays constant.

@@

There is a family of methods for inverting operators with small singular values, known as regularisation methods.  A very useful idea is due to Tychonov.  The problem
$$\Gx f=|ψ〉$$
can be regarded as a least squares problem, where $|ψ〉$ is chosen to minimise the residual $‖\Gx f-|ψ〉‖²$.  The exact solution has residual zero, of course.  The problem is that it is swamped by components with small singular values of $\Gx$.  Suppose we modify this problem, and seek $|ψ〉$ to minimise $‖\Gx f-|ψ〉‖²+λ‖f‖²$, where $λ$ is a positive number, known as a Lagrange parameter for reasons that will become clear shortly.  There are two things to note here.  Firstly, this is still a least squares problem: it is equivalent to minimising 
$$\left‖ \pmatrix{\Gx \cr λ} f-\pmatrix{|ψ〉\cr 0}\right‖².$$
Therefore, it can be solved stably by the usual least squares methods.  Secondly, suppose $f_λ$ is the solution for a given value of $λ$.  If another function, $g$, had the same norm but a smaller residual, then $f$ would not minimise the Tychonov expression.  The Tychonov method is equivalent to constraining the norm of $f$, and seeking the solution with smallest residual given the constraint.  Or, equivalently, constraining the residual and seeking the solution with smallest norm.  This equivalence to a constrained optimisation problem is why $λ$ is a Lagrange parameter.

@@

\topinsert\XeTeXpicfile lcurve-1.pdf \endinsert

An example for a typical application is shown below.  The context is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.  The real state is $x₀$, and it has generated observations $b=Ax₀+e$, where $e$ consists of measurement error.  The problem $Ax=b$ is solved for $x$, with various Lagrange parameters.  At the top left, the norm of $x$ is constrained to be zero, so the solution is $x=0$, and the residual is the norm of $b$.  As the constraint on the norm is relaxed, the solution becomes closer to $x₀$, and the residual decreases.  At the elbow of the curve, the norm of the solution is constrained to be close to the norm of the real state $x₀$, the solution is close to $x₀$, and the residual is the norm of the measurement error.  Relaxing the constraint further allows solutions $x$ that fit the measurements better than the real state $x₀$ does.  Typically, they won't fit much better, so the residual doesn't decrease much below $‖e‖$.  However, the solution now includes components belonging to small singular values, and the norm of the solution will increase as far as it is allowed to.

@@

I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.  However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.  This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
$$f(t)=\int F(ω)e^{iωt}\,dω.$$
This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.  (What is the SVD of the DFT matrix?  It's ${\rm anything}×1×{\rm FT anything}$: Rayleigh theorem.)

A numerical example is shown below.  The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.  A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $|g(a*)|²$, where $g=G⁺|ψ〉$.  These are plotted below, as circles in phase space.  For smaller ensembles, the results were not consistent between draws.

@@

It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the l-curve experiments for the expansion problem.  To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.  In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around $π^½$.  And we're back to the other open question, how to relate the norm of a function to the norm of a set of samples at irregular points.  How can I estimate how much of phase space “belongs” to each amplitude sample, in order to get a volume over which the function has that value?

The obvious way would be to assign amplitudes to the nearest sample.  That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.  What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?  There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.  Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.  I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.  What's the best exponent?  I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.

@@

\topinsert\XeTeXpicfile cats.pdf width \hsize\endinsert

The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉f=〈A|ψ〉$, where $f$ is a vector of weights such that $|A〉f$ approximates $|ψ〉$.  Below it is a plot of the expansions over number states of the singular kets $〈n|u_r〉$, where $|A〉=|U〉SV$.  For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.  The components of the cat have coherent amplitude 2, so their largest number component is $n=4$, and the components drop quickly as $n$ increases further.  Above $r=15$, the singular ket $|u_r〉$ has very small components on these number states, so it hardly overlaps $|ψ〉$.  At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.  In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied.  Seeing that $G$ is perfectly conditioned, this is a very bad discretisation.

The final graph is an L curve when the discrete problem is Tychonov regularised.  It shows the expected shape, with minimum residual around $2×10^{-8}$.  Six regularised solutions are shown above; the modulus of $f_r$ is the disk's saturation, and the phase determines its hue.  Note that least norm regularisation has the desirable property that adjacent $α_r$ have similar $f_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.

\topinsert\XeTeXpicfile catpicard.pdf width \hsize\endinsert

@@

I'm getting a picture of how the final algorithm will look.  At each timestep, the directions belonging to large singular values of $|Dψ〉$ will be used for variational dynamics.  This can be done by solving at Tychonov conditioned problem, for example.  At the same time, the directions with small singular values will be used to keep the discretisation stable.  I think this requires two things: that $|Dψ〉$ is reasonably well conditioned, and that the angle between $H|ψ〉$ and the span of $|Dψ〉$ is small.  This raises a question: are there singular directions that have little effect on $|ψ〉$, but a large effect on $|Dψ〉$?  This seems likely; for example, shuffling the amplitudes to move two components further apart in phase space, then adjusting the weights so that $|ψ〉$ stays the same.  This could be done by a kind of Tychonov anti-regularisation: we improve the stability as much as we can, while constraining the change in $|ψ〉$.  I might need to constrain the norm of $dz$ too, so that linear approximations hold.

@@

\centerline{\XeTeXpicfile psfa.pdf width \hsize}

\centerline{\XeTeXpicfile psv.pdf width \hsize}

@@

@@

The norm of $Gg$ is 1, so this suggests the singular values of $G$ are $π^{-½}$.  This can also be seen from the expansions of coherent states.  The normalised entire expansion of $|α〉$ is
$$g(β*)=〈β|α〉=e^{-½|α|²-½|β|²+β*α},$$
whence
$$‖g‖²=\int |g(β)|²\,d²β=\int e^{-|β-α|²}\,d²β=π.$$
The norm of $|α〉$ is of course 1.  If we used the Bargmann space, we could easily absorb the $π$ into the inner product, and make $G$ a generalised projector.

The lemma gives a very simple singular value expansion of $G$.  Because there is only one nonzero singular value, any orthonormal basis of ket space will do as the left singular vectors.  Let's choose the number states for familiarity.  Then
$$Gg=∑_n |n〉π\int ν_n*(α)g(α)\,d²α=|N〉π\int ν*(α)g(α),$$
where $|N〉=\bigl(|0〉\ |1〉\ |2〉\ …\bigr)$ and $ν(α)=\bigl(ν₀(α)\ ν₁(α)\ …\bigr)$ could be thought of as infinite matrices whose columns are the left singular vectors $|n〉$ and their normalised entire expansions, the right singular vectors $v_n$.  The pesudoinverse, 
$$G⁺|ψ〉=νπ〈N|ψ〉= ∑_n ν_nπ〈n|ψ〉,$$
maps $|ψ〉$ to its normalised entire expansion.  Expanding over number states is useful because the maximum number in a system is usually limited by conservation of energy.  In real systems, the components decay exponentially with $n$, and become smaller than numerical rounding errors at reasonably small numbers.

Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.  (And they do: when I plotted their overlaps with coherent states, they formed rings without support at the origin.)  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.

@@

So what are the critera for adjusting the grid?  I think there are two obvious ones: firstly, that the inverse problem stays well-posed.  Otherwise, we're wasting grid points.  Secondly, when we're expanding a derivative over the tangent space $|Dψ〉$, we want to tilt the tangent space closer to the derivative.  That's a second order effect, and won't come naturally from a least squares adjustment of $f$ and $α$.  The first criterion seems redundant: if we could better represent the goal by shifting two sample points apart, the second criterion will do that automatically.  However, if we only care about the tangent space and not the derivative map, we might need very large adjustments of the amplitudes and weights to get to the point in the tangent space that we want.  So a condition number criterion seems sensible.  There might be other criteria that I've overlooked.

@@

\centerline{\XeTeXpicfile expens.pdf width \hsize}

@@

\centerline{\XeTeXpicfile exp.pdf width \hsize}

@@

\centerline{\XeTeXpicfile spsns.pdf width \hsize}
Amplitudes of components

\centerline{\XeTeXpicfile traj.pdf width \hsize}
Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted)

\centerline{\XeTeXpicfile picard.pdf width \hsize}
Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied.

\centerline{\XeTeXpicfile svnum.pdf width \hsize}
Expected particle number for left singular vectors.  The svs with small sws have large particle number.  When $H$ is quartic, these have large components in $H|ψ〉$.  In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$.

@@

\centerline{\XeTeXpicfile flwc0n10.pdf width \hsize}

\centerline{\XeTeXpicfile flwc0n5.pdf width \hsize}

@@

The things I've noticed so far are:
\item{1.} When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.  It isn't accurate, though.

>>>>

\nonstopmode
\input a4
\input xpmath
\input respnotes \tenrm

@@

Bargmann showed that the Hilbert space of wavefunctions is isomorphic to the space of phase space distributions, under the isometry $〈Γ⁺|$, where
$$〈 Γ⁺|ψ〉(α)=π^{-½}〈 α|ψ〉.$$
Being a linear operator from quantum states to phase space functions, $〈Γ⁺|$ is written as a ket.  It is a pseudoinverse of $|Γ〉$, in the sense that $|Γ〉〈 Γ⁺|$ is the identity operator in state space.  The other composition, $〈 Γ⁺|Γ〉$, projects the functions of a complex variable onto the entire functions, orthogonally under the Bargmann inner product.  (Prove that, or find Bargmann's proof.)  It follows that $f=〈 Γ⁺|ψ〉$ is the function with least norm that satisfies $|Γ〉 f=|ψ〉$, with $‖f‖=\bigl‖ |ψ〉\bigr‖$.

If, following Bargmann, we identify state space with the entire functions, $|Γ〉$ is identified with the projector $〈 Γ⁺|Γ〉$.  Therefore, although $|Γ〉$ is singular, the singularity is easily resolved.  If we constrain the norm of $f$ to be not much greater than the norm of $|ψ〉$, which can be done simply enough by Tychonov regularisation, we ensure that $f$ is close to $〈 Γ⁺|ψ〉$.  And $〈 Γ⁺|$, an isometry, is as nice a function as could possibly be desired.  So it is curious that the discrete version of this problem has caused so much trouble.

@@

The part of $|Γ〉$ that projects functions onto entire functions can be written as follows
$$π^{-½}〈β|Γ〉f=π^{-1}\int 〈β|α〉e^{-½|α|²}f(α*)\,d²α=π^{-1}\int 〈0|D(α-β)|0〉e^{-½|α|²}f(α*)\,d²α.$$
This appears to be a convolution of the function $e^{-½|α|²}f(α*)$ with the kernel $〈0|D(α)|0〉$.  There are many theorems to the effect that real convolution operators are badly conditioned.  It is interesting that these results do not apply to the complex convolution $|Γ〉$: as a projector, it is as nice an operator as it could possibly be.  I think the loophole is that of real functions can only oscillate at a limited rate, so that their Fourier transforms have finite norm.  Complex functions have space in the Argand plane to oscillate arbitrarily rapidy.  And the kernel $〈0|D(α)|0〉$ does: it is a wave packet with carrier frequency proportional to the imaginary part of $α$.

@@

The operator $|Γ〉$ acts on functions defined over the whole complex plane.  For states with high particle number, the smallest preimages have support far from the origin.  We can not compute with such functions: any discretisation will sample values of $f$ on a finite subset of the plane.  Unfortunately, the conditioning of these restricted operators will be worse than that of $|Γ〉$.  We have excluded the small preimages of the states with large particle number, so they must be represented by functions with support close to the origin, but with large norms.  Hence the restricted operators will be badly conditioned.

@@

The sizes of states and their expansions can be read from the singular value expansion of $\Gx $.   If the domain and range of $\G$ were the same space, it would have an eigenvalue decomposition; the SVE is a generalisation of that.  It has the form
$$\Gx f = ∑_{i=1}^∞ |u_i〉σ_i〈v_i,f〉.$$
The kets $|u_i〉$ are an orthonormal basis of state space, known as the left singular vectors of $\Gx$.  The $σ_i$ are positive real numbers, called singular values.  The functions $v_i$ are are orthonormal under the Bargmann inner product, and are called right singular vectors.

In the case where $χ(z)=1$, and $\Gx=\G$, the expansion is quite arbitrary.  The left singular kets can be any orthonormal basis, the right singular vector $v_i$ is the unique entire function for which $\G v_i=|u_i〉$, and every singular value is 1.  If the domain and range of $\G$ were the same, this would indicate that it was a projector.  The component of $|ψ〉$ in the null space of $\G$ is discarded, being orthogonal to every $v_i$, while the other component retains its expansion coefficients as the basis $v_i$ is replaced by $|u_i〉$.

When the operator is restricted, the SVE gives more information. If $\Gx$ were one to one, each $v_i$ would be the unique preimage of $|g_i〉$; since $\Gx$ is singular, $v_i$ is the preimage with least norm.  The singular values are the ratios of $‖\Gx v_i‖=‖σ_i|u_i〉‖$ to $‖v_i‖$.  If $|u_i〉$ has a low particle number, it can be expanded efficiently: $v_i$ will be close to $〈Γ⁺|u_i〉$, and $σ_i$ will be close to 1.  If $|u_i〉$ includes components with large particle number, its expansion will have a larger norm than itself.  In this case, $v_i$ will be a truncation of $〈Γ⁺|u_i〉$ to small amplitudes, multiplied by a large constant to maintain normalisation, and $σ_i$ will be small.

Before we study the singular values and vectors of $\Gx$, consider why small singular values are a problem.  The pseudoinverse of $\Gx$ with least norm is
$$〈Γ⁺_χ|=∑_{i=1}^∞ v_iσ_i^{-1}〈u_i|.$$  Suppose this is applied to a state $|ψ〉$ that has a low expected particle number, but small components with large particle number.  Such a component will overlap some $|u_i〉$ belonging to small $σ_i$, and $〈Γ⁺_χ|$ will multiply it by $σ_i^{-1}$, generating large components of those $v_i$ in the expansion function.  In the continuous expansion, this isn't a problem: the information is still there, and the state can be recovered by applying $\Gx$ to the expansion.  When the expansion is discretised, however, the small, low particle number component of $〈Γ⁺_χ|ψ〉$ that contain the important information about $|ψ〉$ are very likely to be lost in the large components of $〈Γ⁺_χ|ψ〉$ that give minor details of high particle number components of $|ψ〉$.

Let's compute some singular values and vectors.  (And the the wrong answer—try again.)  Suppose that $χ$ has circular symmetry, with $χ(z)=χ(|z|)$.  We can rewrite this
$$\eqalign{\Gx f&=π^{-½}∑_n(n!)^{-½}|n〉\int αⁿe^{-|α|²}χ(α*)f(α*)\,d²α\cr
	&=π^{-½}∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)\int_0^{2π}f(re^{-iθ})e^{inθ}\,dθdr\cr
	&=2\sqrt π∑_n(n!)^{-½}|n〉\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr,}$$
where
$$f(re^{iθ})=∑_{m=-∞}^∞ f_m(r)e^{imθ}.$$
Presumably, the right singular vectors have the $f_n$ of least norm such that 
$$2\sqrt π(n!)^{-½}\int_0^∞ r^{n+1}e^{-r²}χ(r)f_n(r)dr=1.$$

@@

\topinsert \hbox{\XeTeXpicfile rsv2.pdf width 0.5\hsize
\XeTeXpicfile rsv3.pdf width 0.5\hsize}
\XeTeXpicfile rsv4.pdf width 0.5\hsize\hfil\break
The right singular functions $v_i$ for the truncated expansion operator $\Gx$, where phase spaces is truncated to a square about the origin, whose size varies.  These are functions of a complex variable: the brightness of the plot indicates modulus, and the color indicates phase.\endinsert

@@

The most basic issue is the Picard condition.  This restates a simple observation: for the problem $\G \dot f=H|ψ〉$ to have a solution $\dot f$, the expansion of $\dot f$ over an orthonormal basis must have a finite norm.  If the basis is chosen as the right singular vectors $v_i$, this means that the sum of the squared moduli of their coefficients must be finite.  The component of $v_i$ is $〈u_i|H|ψ〉/σ_i$.  So the Picard condition is that, for the equation to have a solution, the series
$$∑_{i=1}^∞{\bigl|〈g_i|H|ψ〉\bigr|²\over σ_i²}$$
must converge.  The singular values reduce with $j$, by definition; this requires that the components of the left singular vectors reduce somewhat faster.

The variational problem seems to be pretty well-posed: if the initial superposition is the phase space function sampled on a grid, then the state that comes from a variational change in the amplitudes matches $H|ψ〉$ very closely.

@@

There is a family of methods for inverting operators with small singular values, known as regularisation methods.  A very useful idea is due to Tychonov.  The problem
$$\Gx f=|ψ〉$$
can be regarded as a least squares problem, where $|ψ〉$ is chosen to minimise the residual $‖\Gx f-|ψ〉‖²$.  The exact solution has residual zero, of course.  The problem is that it is dominated by components belonging to singular values of $\Gx$.  Suppose we modify this problem, and seek $|ψ〉$ to minimise $‖\Gx f-|ψ〉‖²+λ‖f‖²$, where $λ$ is a positive number, known as a Lagrange parameter for reasons that will become clear shortly.  There are two things to note here.  Firstly, this is still a least squares problem: it is equivalent to minimising 
$$\left‖ \pmatrix{\Gx \cr λ} f-\pmatrix{|ψ〉\cr 0}\right‖².$$
Therefore, it can be solved stably by the usual least squares methods.  Secondly, suppose $f_λ$ is the solution for a given value of $λ$.  If another function, $g$, had the same norm but a smaller residual, then $f$ would not minimise the Tychonov expression.  The Tychonov method is equivalent to constraining the norm of $f$, and seeking the solution with smallest residual given the constraint.  Or, equivalently, constraining the residual and seeking the solution with smallest norm.  This equivalence to a constrained optimisation problem is why $λ$ is a Lagrange parameter.

@@

\topinsert\XeTeXpicfile lcurve-1.pdf\vskip 5mm\noindent\it L curve for a typical application of Tychonov conditioning\endinsert

An example for a typical application is shown above.  The context in which Tychonov regularisation is usually applied is that some observations $b$ are predicted by a physical state $x$, according to the linear law $Ax=b$.  The real state is $x₀$, and the observed data are $b=Ax₀+e$, where $e$ consists of measurement error.  The problem $Ax=b$ is solved for $x$, and regularised with various Lagrange parameters.  At the top left, the norm of $x$ is constrained to be zero, so the solution is $x=0$, and the residual is the norm of $b$.  As the constraint on the norm is relaxed, the solution becomes closer to $x₀$, and the residual decreases.  At the elbow of the curve, the norm of the solution is constrained to be close to the norm of the real state $x₀$, the solution is close to $x₀$, and the residual is the norm of the measurement error.  Relaxing the constraint further allows spurious solutions $x$ that fit the measurements better than the real state $x₀$ does.  Typically, they won't fit much better, so the residual doesn't decrease much below $‖e‖$.  However, the solution now includes components belonging to small singular values, and the norm of the solution will increase as far as it is allowed to.

@@

I've been assuming that the variational problem is discrete, because it comes from a discretisation of an integral equation.  However, the integral equation is a complex convolution, and for some values of the complex amplitude, the kernel oscillates arbitrarily sharply.  This is the same sort of integral equation as the Fourier transform, where we know $f(t)$, and want to find $F(ω)$ such that
$$f(t)=\int F(ω)e^{iωt}\,dω.$$
This also has a kernel that oscillates at a rate proportional to $ω$, and it also has very stable solutions.

A numerical example is shown below.  The state $|ψ〉$ is the Schrödinger cat that a coherent state becomes after half a cycle in a quartic oscillator; the amplitude of the initial coherent state was $2$.  A set of 40 sample amplitudes $α_r$ was drawn from the probability distribution $P(α)=|e^{-½|α|²}〈G⁺|ψ〉(α*)|²$.  These are plotted below, as circles in phase space.  For ensembles with fewer components, the results were very inconsistent between draws.

The state $|ψ〉$ was expanded in the form $|ψ〉=|A〉\vec f$, where $|A〉=\pmatrix{|α₁〉&⋯&|α_R〉}$, and $\vec f$ is a column vector of $R$ complex weights.  This was done with Tychonov regularisation, for various Lagrange parameters: these and the residuals are shown with the graphs.  The complex weights $c$ are indicated by brightness and color as above.  Note that least norm regularisation has the desirable property that nearby $α_r$ have similar $c_r$: the least norm solution to $a+b+c=X$ is $a=b=c=X/3$.

\topinsert\XeTeXpicfile cats.pdf width \hsize
\vskip 5mm\it A cat state expanded over a random set of 40 amplitudes, as described in the text.\endinsert

@@

It seems that, when two amplitudes are very close, the solutions where the weights cancel out are better fits than the smooth ones—that seems to be the case in the L-curve experiments for the expansion problem.  To avoid the amplitudes clustering, I need to find out how good a fit is possible with evenly spaced ampltiudes, and set the regularisation parameter so that the residual is on that order.  In fact, I already know how large the phase space expansion should be: the function being sampled should have a norm around 1.

@@

\topinsert\XeTeXpicfile catpicard.pdf width \hsize\vskip 5mm\it
Picard plot for cat expansion. \endinsert

The next Figure is a Picard plot for the Galerkin discretisation $〈A|A〉\vec f=〈A|ψ〉$, where $\hat f$ is a vector of weights such that $|A〉\vec f$ approximates $|ψ〉$.  The singular value decomposition is $|A〉=|U〉SV$.  For the kets to be orthogonal, $|u_r〉$ must have support close to $n=r$.  The components of the cat have coherent amplitude 2, so their largest number component is $n=4$, and the components drop quickly as $n$ increases further.  Above $r=15$, the singular ket $|u_r〉$ has very small components on these number states, so it hardly overlaps $|ψ〉$.  At the same time, the singular kets start to have particle numbers too large to be consistent with the sample amplitudes $α_r$, so the singular values decrease.  This is shown in the plot of Fock state components of the singular kets.  In fact, they decrease more rapidly than the components of $|ψ〉$, and the discrete Picard condition is not satisfied: the discrete doesn't have a finite solution.  Seeing that the continuous operator $\G$ is perfectly conditioned, this is a very bad discretisation.

The final graph is an L curve when the discrete problem is Tychonov regularised.  It shows the expected shape, with minimum residual around $2×10^{-8}$.

@@

I'm getting a picture of how the final algorithm will look.  At each timestep, the directions belonging to large singular values of $|Dψ〉$ will be used for variational dynamics.  This can be done by solving at Tychonov conditioned problem, for example.  At the same time, the directions with small singular values will be used to keep the discretisation stable.

@@

\topinsert 
\vskip -40mm
\XeTeXpicfile psfa.pdf width \hsize
\vskip -90mm
\XeTeXpicfile psv.pdf width \hsize\endinsert

@@

An obvious way would be to assign amplitudes to the nearest sample.  That makes the volume for each sample a quite irregular prism; maybe there is a simple way to calculate the volumes of those, but I don't know it.  What if we assign a value to each point as an average over the sampled values, weighted, say, in proportion to the squared distances?  There would need to be some extra condition to ensure the function tends to zero at points remote from the samples.  Something like $x^{-2}e^{-|x|²}$ sounds plausible: it's quadratic at sort range, where the gaussian is flat, but outside the support of the samples, it decays in the right way for a phase space expansion.  I think the short range part needs to be a power law, so that it doesn't have a length scale built in when the distances between samples are random.  What's the best exponent?  I expect it depends on the dimension of the phase space, which would determine how many samples we expect to find close to a given one.

@@

Truncating Fock space has a similar effect to truncating the singular values, assuming the singular directions belonging to small values have large particle numbers.  And indeed, the trajectories for the average amplitude change in similar ways due to truncation: they pick up an extra loop in the complex plane.

@@

@@

\topinsert \XeTeXpicfile expens.pdf width \hsize \endinsert

@@

\topinsert \XeTeXpicfile exp.pdf width \hsize \endinsert

@@

\topinsert \XeTeXpicfile spsns.pdf width \hsize
Amplitudes of components \endinsert

\topinsert \XeTeXpicfile traj.pdf width \hsize
Left: expectation value $〈a〉$ at sampled times.  Right: residual of fit by superposition (solid) and condition number of Jacobian of fitted superposition (dotted) \endinsert

\topinsert \XeTeXpicfile picard.pdf width \hsize
Picard plots at fitted times.  Circles are singular values, crosses components of LSVs, red stars components of RSVs.  The discrete Picard condition is clearly not satisfied. \endinsert

\topinsert \XeTeXpicfile svnum.pdf width \hsize
Expected particle number for left singular vectors.  The svs with small sws have large particle number.  When $H$ is quartic, these have large components in $H|ψ〉$.  In order for the svs to be orthogonal, the points are likely to lie around the line ${\rm sv}=n$. \endinsert

@@

\topinsert \XeTeXpicfile flwc0n10.pdf width \hsize \endinsert

\topinsert \XeTeXpicfile flwc0n5.pdf width \hsize \endinsert

@@

The things I've noticed so far are:
When a superposition is iteratively fitted to the exactly propagated state, adding a term to the cost proportional to the Jacobian $|Dψ〉$ keeps the fit stable for an entire cycle.  It isn't accurate, though.
